Spatial Smoothing in Mass Spectrometry Imaging

### 2017-04-20 docs/msci-paper/paper_ssmsi.tex
The final project submission.

% - intro: Metabolomics
% - research problem: Spatial smoothing
% - prelims: Design an appropriate topic model
% - method: Provide the methodology topic model
% - experiments: synthetic data
\begin{abstract}
In this paper, we target a data modelling approach used in computational metabolomics; to be specific, we assess whether spatial smoothing improves the topic term and noise identification. By assessing mass spectrometry imaging data, we design an enhancement for latent Dirichlet allocation-based topic models. For both data pre-processing and model design, we survey relevant research. Further, we focus on presenting our methodology in detail providing the preliminaries to the basic models and guiding through the performed adjustments in the proposed model variations. To assess the application's performance, we run the model comparison on a number of diverse synthetic datasets.
\end{abstract}

@@

% Purpose:
% - A spatial smoothing application in visual data using unsupervised ML
% - Focus: Metabolomics and topic modelling
\par In this research paper, we assess an application of spatial smoothing in visual data; that is, we induce continuity among data elements. The spatial smoothing application is particularly targeted to be applied for unsupervised pattern recognition. To be more specific, our focus is to model a biomedical application in the field of metabolomics. Furthermore, we limit the scope of the applied unsupervised machine learning techniques to the branch of topic modelling.

% Basis:
% - Visual data: MSI application in metabolomics
% - MSI data structure
\par The characteristics of our utilised metabolomics data are expressed in the form of mass spectrometry imaging (MSI). Effectively, we use MSI to visualise the metabolomics data in the form of spatial distribution. Speaking of the metabolomics data, it contain information about ionised metabolites. Note that metabolites are molecules produced by the chemical process of metabolism, whereas by ionisation we refer to the information capture method. In other words, MSI data is a visualisation of ion (sampled metabolite) distributions: the complete dataset is the whole image; the image's pixel is a particular sampling region; and each region contains intensities of ions with unique mass-to-charge $m/z$ values.

% Basis:
% - Unsupervised machine learning: Topic modelling
% - topic modelling w.r.t. MSI
\par Speaking of our machine learning application, topic modelling is a technique used to infer unlabelled topic distributions based on data's underlying semantic structure. With respect to MSI data, we can model the topic distributions over an image and its every pixel; furthermore, topic modelling can express the types of ions corresponding to particular topics. Since we use a topic model to model metabolomics data in a statistic manner, the utilised topic models are tuned to reflect the metabolomics environment as realistically as possible. Relating to our research targets, spatial smoothing is one of such environment settings.

% Research Problem:
% - Noisiness
% -- Fragmentation
% -- Retention time
% - Loss of information
% -- Overlapping topics
\par The basis of the project's research problems comes from the limitations of current metabolite sampling techniques. The main limitation is the loss of information caused by the metabolite ionisation. As a consequence, the MSI data is noisy. To expand on the noisiness, it is caused by metabolite fragmentation and the limitation to tune different metabolite retention times. By metabolite fragmentation, we refer to a metabolite split; the split would cause the captured ions to possess unexpected values. Speaking of the retention time, the intensity value of each ion type varies with respect to time; therefore, since each ion is captured at a different state of its retention, each ion type would possess some variance with respect to its intensity. Finally, note that the loss of information might be also caused by overlapping ion topics. Since different ion topics can contain same ion types, the topic possessing a lower ion intensity value would be overwhelmed and, thus, not reflected in the MSI data.
% Contributions:
% - A specific and extensive assessment of the SS application
% - A tuned topic model and well-organised experiment settings
\par In this paper, we contribute to the research in MSI by carrying an extensive assessment of the spatial smoothing application. The assessment is carried in both quantitative and qualitative manners: we assess the performance on a number of diverse datasets; also, our experiments are designed to reflect the nature of the MSI data reflecting computational metabolomics. Furthermore, we provide a Python implementation of a tuned topic model; also, we establish maintainable experiment settings. By the tuned topic model, we mean that the model's implementation is particularly designed to meet the characteristics of the MSI data. Speaking of the experiment settings, note that we are carrying the experiments using Jupyter notebooks. Effectively, the use of the notebooks creates a portable and well-documented environment to initialise the experiment settings and execute the topic modelling. As a result, external parties could run the released notebooks and reproduce the experiment results in a swift manner.

@@

% LDA Intro:
% - Hierarchical structure
% - Generative model
% - Bayesian methods
% - Assumptions
\par The research project targets a specific branch of topic models. The branch consists of Latent Dirichlet Allocation (LDA) derivatives. Note that the initial LDA model was introduced by Blei et al. \cite{blei2003latent}. One of the model's key characteristics is the three-level hierarchical treatment of the data. In the context of the utilised MSI data, the hierarchical structure can be perceived as follows: in the highest level, we have an MSI image; in the middle level, we have a pixel of an MSI image; and in the lowest level, we have the intensities of particular ions in a pixel.

% Generatitive model:
% - Generative process
\par Another model's key characteristic is the generative treatment of the data. By a generative model, we mean that the latent data instances are treated as a result of a mixture of underlying parameters drawn from probability distributions. In other words, the generative data treatment induces randomness in the end products of the data; however, note that the source of the data -- the lowest level parameters of the probability distributions -- remain the same. The key aspect of the generative model is the degree of freedom in the connections of random variables; this notion allows modelling more realistic, thus, more complex data relations. Ultimately, the rationale of the generative model is based on recovering the underlying semantic structure. Effectively, in order to recover the generative process of the data, we delve into the applications of Bayesian methods.

% Bayesian methods
% - Inference: sampling
% - Connection to LDA
\par By applying Bayes' rule, we can express the underlying semantic structure in the form of a posterior probability distribution. However, note that we can not analytically compute such posterior expression. Instead, we can estimate the posterior distributions using optimisation or direct sampling. In this project, we particularly focus on the underlying semantic structure's inference using direct sampling. Speaking of the sampling-based inference method applications to LDA-like models, the ground-work has been established by Griffiths and Steyvers \cite{griffiths2004finding}. The authors have utilised the collapsed Gibbs sampling -- a Markov chain Monte Carlo (MCMC) algorithm. Effectively, the method integrates the uncertainty out; so that we sample the entities of interest directly. More specific details about the rationale of the collapsed Gibbs sampling application will be provided in Section 4.

% Assumptions
% - Exchengeability (BoW)
% - Discrete data
\par Relating to the initial LDA model, the authors have set the following assumptions for the utilised data: exchangeability among the inner components of the lower and middle data hierarchy levels; also, a discrete treatment of the lower-level data. By exchangeability, it is meant that the components follow the bag-of-words principle; that is, the order of the data has no correlation with the underlying semantic structure. Speaking of the discrete data treatment, it is assumed that the lower-level components have no spatial connection. With respect to our project, both assumption types -- exchangeability and discreteness -- do not correspond to the characteristics of the MSI data. Therefore, we will establish a methodology to relax the latter assumptions.

@@

% Terminology intro:
% - Introduce the three-level hierarchical structure
\par In this subsection, we introduce our topic modelling terminology. In order to put the MSI data in the topic modelling context, the components of  LDA's three-level hierarchical structure are defined as follows: \textit{a corpus} is the whole image of a sample; \textit{a document} is an MSI image's pixel; and \textit{a word} is an interval of mass-to-charge values corresponding to a particular ion. Starting from now on, we will use the latter topic modelling concepts to introduce the LDA model's architecture and notation.

% LDA design
% - Graphical model
\par Since LDA-like models are also graphical models, the dependence of random variables can be illustrated using graphs. In order to familiarise with the architecture and the variables of LDA-like models, we provide Figure \ref{fig:arch_LDA-init} illustrating the initial LDA model in the plate notation.
\begin{figure}[h]
  \centering
  \includegraphics[width=0.25\textwidth]{LDA-initial.png}
  \caption{The initial LDA model's architecture.}
  \label{fig:arch_LDA-init}
\end{figure}
The circles indicate the model's variables: the coloured circle corresponds to the observable variable, whereas the uncoloured circles correspond to the hidden variables. Effectively, the hidden variables define the model's underlying structure. Speaking of the plates, the plate denoted by $K$ corresponds to the number of topics; the plate denoted by $T$ corresponds to the number of documents, and the plate denoted by $N$ corresponds to the number of words. Effectively, the letters in the bottom right corners indicate the total number of variables. Therefore, a corpus has a $T$ number of documents, and each document has an $N$ number of words.

% Terminology
% - Parameter meaning
% - Listing
\par Before providing a listing with the definitions of the LDA variables, we introduce the purpose of the variables. The variables denoted by $\theta$ and $\phi$ correspond to the underlying probability distributions (e.g., in the case of the initial LDA model, we use Dirichlet distributions). It follows that the variables denoted by $\alpha$ and $\beta$ act as the parameters of the latter probability distributions; note that in the context of machine learning, such auxiliary parameters are called hyper-parameters. Effectively, \textit{a hyper-parameter} allows tuning a machine learning model for a particular dataset application. As a side reference, note that by \textit{a vocabulary} we refer to a collection of terms reflecting a fixed range of the words. At this point, we provide the following list containing the definitions of the initial LDA model's variables: 
\begin{itemize}
	\item $K$ is the number of topics;
	\item $T$ is the number of documents;
	\item $N$ is the number of words per document;
	\item $V$ is the size of a vocabulary;
	\item $w$ is a word;
	\item $z$ is a word's topic assignment;
	\item $\theta_t$ is the topic distribution over the document $t$;
	\item $\phi_k$ is the vocabulary term distribution over the topic $k$;
	\item $\alpha$ is the hyper-parameter for the topic distributions;
	\item $\beta$ is the hyper-parameter for the vocabulary term distributions.
\end{itemize}

@@

% MSI dataset intro:
% - Introduce the raw data format
% - Describe the data pre-processing requirements
\par In this subsection, we introduce the qualities of the raw MSI data. Furthermore, we set the requirements for the raw data's pre-processing; note that the pre-processing serves as an auxiliary method making the raw MSI data compatible for a scalable topic modelling application. As a side note, the basis of our applied MSI data characteristics is established from the mzML data format. Conveniently, we parse mzML data using the \texttt{pymzml} Python library.

% Raw data format
% - Data meaning
% - Explain the raw data by looking into it
\par Essentially, raw MSI data contain mass spectra of a sample -- the sample's every pixel is expressed in the form of a mass spectrum. By a mass spectrum, we refer to a map from a mass-to-charge value to an intensity value. Ultimately, we can establish a continuous notion of the MSI data by ordering the mass-to-charge values. Note that since the sampling equipment can detect the mass-to-charge values in the millidalton (mDa) precision, the MSI data is sparse (i.e., a large portion of mass-to-charge values are mapped to the zero intensity).

% Data pre-processing requirements
% - Vocabulary generation
% - Intensity normalisation
% - Dismissing inferior information
\par In the context of topic modelling, the raw MSI data possess a large vocabulary (above $5000$ terms). Note that we consider every mass-to-charge value as a word; whereas every intensity value is perceived as a word's occurrence count. Further, since the intensity values could spike up above $1000$, the time complexity of the latent topic inference would require pro-longed runs. To overcome the introduced scalability issues, we carry data pre-processing -- the applied techniques are discussed in the upcoming methodology section.

@@

% Contributions
% - Naive AR model application
% - Motivating for the state-of-the-art AR model application
\par The impact of proving the hypothesis would bring the following contributions:
\begin{itemize}
	\item Improving the detection of overlapping topics and vocabulary terms;
	\item Reducing the noisiness of the MSI data;
	\item Motivating a further research in the spatial smoothing application in MSI data.
\end{itemize} 
To expand on the overlapping topic detection, the issue arises when two underlying topics are made of similar vocabulary terms: instead of a separate representation, the topics are merged. We intend to identify the flow of distinct topics by the application of spatial smoothing; as a consequence, the spatial smoothing would also impact the data noisiness reduction. Ultimately, if a naive spatial smoothing application displayed a performance improvement, the contribution would set a basis to utilise state-of-the-art auto-regression techniques on the MSI data.

% Approach
% - Clearly define the experiment settings
% - Tune the models by studying synthetic datasets.
\par To my knowledge, the impact of the spatial smoothing application to the MSI domain has not yet been thoroughly studied. For the latter reason, this research project serves as an exploratory assessment of the spatial smoothing application: we introduce the rationale behind the applied methodology; also, we clearly define the range of the experiment settings. To give a brief intuition about the methodology, the study assesses the domain-specific parameter tuning and its impact on a diverse range of synthetic datasets.

@@

% Intro (separate into three)
% - MSI DATA:
% -- alonso2015analytical
% -- Review the metabolomics feature pre-processing
% -- Review the metabolomics alternative approaches
% - Review the implementation approaches
% -- blei2012probabilistic
% -- Gibbs sampling
% -- DTM
\par In this section, we review the ground-work carried on the following aspects: the pre-processing of MSI data; the rationale of the utilised topic models; and novel approaches exploiting the characteristics of MSI data. Note that the covered groundwork is selected to reflect the rationale of the utilised techniques. In order to establish the basis of the state-of-the-art computational metabolomics methodology, we consult the survey carried by Alonso et al. \cite{alonso2015analytical}; speaking of the key probabilistic topic modelling branches, we consult the survey carried by Blei \cite{blei2012probabilistic}.

@@

\par In order to establish a scalable topic inference, we review the following MSI data pre-processing techniques: data normalisation; feature binning; and noise reduction. To start with, data normalisation serves as a method establishing an adaptable data structure. Bolstat et al. \cite{bolstad2003comparison} have proposed a data normalisation method called linear baseline scaling. Note that the method is particularly targeted at sparse datasets. Effectively, the method is applicable to numerical features; note that the method's work-flow is carried as follows: we find the largest numerical feature of all data instances; then, we calculate the scaling factors by aligning the largest values to a pre-set upper threshold; finally, we align the remaining numerical features based on the established scaling factors. Relating linear baseline scaling to applications on MSI data, Kohl et al. \cite{kohl2012state} have shown that the method's application does not produce a significant loss of information to establish a well-performing MSI data inference.

% - Binning
% -- Evenly spaced 
% -- craig2006scaling
% -- Spectral peaks
% -- de2008nmr
\par Speaking of feature binning, the technique is used to merge the instances of a feature space; as a result, feature binning reduces the data dimensionality. Note that by the MSI data feature space, we refer to distinct $m/z$ values. Since raw MSI data is sparse, a successful application of feature binning is based on identifying appropriate $m/z$ boundaries reflecting unique metabolite types. To introduce some examples, the research carried by Craig et al. \cite{craig2006scaling} have utilised an equally spaced feature binning. Even though the authors have succeeded to reduce the data dimensionality, they have encountered a loss of information by splitting the metabolite topics into arbitrary regions. As an alternative approach, De et al. \cite{de2008nmr} have performed a feature binning based on the spectral peak identification. Effectively, the technique induces a dynamic notion of bin boundaries which are based on the identified intensity peak regions. As a result, the merged $m/z$ values serve as a better representation of metabolite terms.

% - Denoising (removing small intensities)
% -- smith2014proteomics
\par The last reviewed MSI data pre-processing techniques addresses the MSI data noisiness. Even though the MSI data noise reduction is an open research problem going well beyond data pre-processing, Smith et al. \cite{smith2014proteomics} have shown that the application of a general data pre-processing routine displays performance improvements. For example, the authors have induced a lower intensity bound. Effectively, the intensity values below the intensity threshold would be treated as insignificant and/or as a product of data capturing device imperfections. By applying this procedure, the authors have successfully reduced the data dimensionality and, thus, increased the data scalability.

@@

% Intro
% - Inference
% - Derivatives
% - Gibbs sampling
% -- griffiths2004finding
\par In this subsection, we review the key characteristics of the prospective topic modelling inference techniques and model variations. Speaking of the inference techniques, we employ the rationale of a sampling-based LDA model. Griffiths and Steyvers \cite{griffiths2004finding} have displayed a successful application of the collapsed Gibbs sampler for the topic inference of both textual and visual data. To expand on the collapsed sampling, the technique allows skipping the estimation of $\theta$ and $\phi$ values; instead, the inference is based on the notion of the assignment counts. In order to estimate a topic assignment's probability, the authors suggest using the following expression:
\begin{equation*}
P(z_i = k | z_{-i}, w) \propto \dfrac{n_{-i, k}^{(w_i)} + \beta}{n_{-i, k}^{(\cdot)} + V\beta}\cdot \dfrac{n_{-i, k}^{(t_i)} + \alpha}{n_{-i, \cdot}^{(t_i)} + K\alpha}.
\end{equation*}
note that $i$ refers to the current iteration ($-i$ refers to the previous iteration); $k$ refers to a particular topic; and $n$ refers to the count of the instances indicated by the term's superscript. If required, $\theta$ and $\phi$ can be sampled from the following distributions:
\begin{align*}
\phi_{k,w} & \sim \mbox{Dirichlet}\bigg(\dfrac{n_{k}^{(w)} + \beta}{n_{k}^{(\cdot)} + W\beta}\bigg),\\
\theta_{t,k} & \sim \mbox{Dirichlet}\bigg(\dfrac{n_{k}^{(t)} + \alpha}{n^{(t)} + K\alpha}\bigg).
\end{align*}
Essentially, more accurate representations of $\theta$ and $\phi$ are obtained by running the Gibbs sampler until a sufficient convergence: we preserve the $\theta$ and $\phi$ values of each iteration; when the sampling is finished, the average of the preserved values is an accurate approximation of the underlying $\theta$ and $\phi$ values.

>>>>

% - intro: Metabolomics
% - research problem: Spatial smoothing
% - prelims: Design an appropriate topic model
% - method: Provide the methodology topic model
% - experiments: synthetic data

\begin{abstract}
In this paper, we target a data modelling approach used in computational metabolomics; to be specific, we assess whether spatial smoothing improves the topic term and noise identification. By assessing mass spectrometry imaging data, we design an enhancement for latent Dirichlet allocation-based topic models. For both data pre-processing and topic model design, we survey relevant research. Further, we present the proposed methodology in detail providing the preliminaries and guiding through the performed topic model enhancements. To assess the performance, we evaluate the spatial smoothing application on a number of diverse synthetic datasets.
\end{abstract}

@@

% Purpose:
% - A spatial smoothing application in visual data using unsupervised ML
% - Focus: Metabolomics and topic modelling
\par In the research project, we assess an application of spatial smoothing in visual data; that is, we induce continuity among data elements. The spatial smoothing application is particularly targeted to be applied for unsupervised pattern recognition. To be more specific, our focus is to model visual metabolomics data using a particular branch of unsupervised machine learning -- topic modelling.

% Basis:
% - Visual data: MSI application in metabolomics
% - MSI data structure
\par The characteristics of the utilised metabolomics data are expressed in the form of mass spectrometry imaging (MSI). To briefly introduce mass spectrometry, the method captures a biological tissue in the form of mass spectrum: at the start, the metabolites of a biological tissue are ionised (by metabolites, we refer to the tissue's contents -- the product of a metabolism chemical process); after the initial step, a mass spectrometer captures the intensity of each ionised metabolite; as a result, we obtain information about the tissue's contents and their concentrations.

% Basis:
% - MSI data structure
\par Relating the latter procedure to MSI, note that we can partition the tissue into small regions. Effectively, the tissue's partitioning corresponds to a higher granularity of the data. As a result, we can identify the contents of the tissue's particular regions. As an example, we provide Figure \ref{fig:msi} illustrating the rationale behind MSI: the image is the whole tissue; the image's pixel is the tissue's particular region; and each pixel contains the intensity values of ions (a different ion type has a unique mass-to-charge $m/z$ value). In order to visualise MSI data, we would look into a specific mass-to-charge value. Effectively, the intensity of a pixel would be set to the intensity value of the respective ion type.

\begin{figure}[h]
  \centering
  \includegraphics[width=0.42\textwidth]{msi3.png}
  \caption{The MSI data structure.}
  \label{fig:msi}
\end{figure}

% Basis:
% - Unsupervised machine learning: Topic modelling
% - topic modelling w.r.t. MSI
\par Going into the machine learning application, note that we use topic modelling for the inference of topic distributions; effectively, the inferred topic distributions would be treated as the underlying semantic structure of a sampled tissue. Since our data is MSI-based, we model the topic distributions over an image and its every pixel; also, we model the types of ions corresponding to particular topics. Since our machine learning application is based on Bayesian methods, we tune the model parameters to reflect the metabolomics environment as realistically as possible. Ultimately, spatial smoothing is one of such environment settings.

% Research Problem:
% - Noisiness
% -- Fragmentation
% -- Retention time
% - Loss of information
% -- Overlapping topics
\par The basis of the project's research problems comes from the limitations of current metabolite sampling techniques. One of the limitations is the presence of noise. For example, MSI data could be distorted by noise as a result of metabolite fragmentation -- the ionisation would split the metabolite's structure causing a faulty mass-to-charge value capture. Another limitation is the overlap of the molecules with similar structures. Effectively, some of the produced ions could have same mass-to-charge values; as a result, the ion possessing a lower intensity value would be overwhelmed and, thus, not reflected in MSI data. A possible approach mitigating the issue is direct infusion; however, in this project, we investigate whether the latter issue could be mitigated using a computational approach.

% Contributions:
% - A specific and extensive assessment of the SS application
% - A tuned topic model and well-organised experiment settings
\par We contribute to the research in MSI by carrying an extensive assessment of the spatial smoothing application. The assessment is performed in both quantitative and qualitative manners: we assess the performance on a number of diverse datasets; also, our experiments are designed to reflect the environment settings of computational metabolomics. Furthermore, we provide Python implementations of the proposed topic model and the experiment settings. Note that the experiments are performed in Jupyter notebooks; effectively, the practice induces a portable and well-documented environment for initialising and running the experiments. The notebooks are also useful in validating the experiment results: external parties would be able to re-run the experiments in a swift manner.

@@

% LDA Intro:
% - Hierarchical structure
% - Generative model
% - Bayesian methods
% - Assumptions
\par The research project targets a specific branch of topic models. The branch consists of Latent Dirichlet Allocation (LDA) derivatives. Note that the initial LDA model was introduced by Blei et al. \cite{blei2003latent}. One of the model's key characteristics is the three-level hierarchical treatment of the data. In the context of the utilised MSI data, the hierarchical structure can be perceived as follows: in the highest level, we have an MSI image; in the middle level, we have the pixels of an MSI image; and in the lowest level, we have the intensities of particular ions in a pixel.

% Generatitive model:
% - Generative process
\par Another model's key characteristic is the generative treatment of the data. By a generative model, we mean that the latent data instances are treated as a result of a mixture of underlying parameters. In other words, the generative data treatment induces randomness in the end product of the data generation; however, note that the source of the data -- the lowest level parameters -- remain the same. The key aspect of the generative model is the degree of freedom in the connections of random variables; this notion allows modelling more realistic, thus, more complex data settings. Ultimately, the rationale of a generative machine learning model is based on recovering the underlying semantic structure. To do this, we employ the rationale of Bayesian methods.

% Bayesian methods
% - Inference: sampling
% - Connection to LDA
\par By applying Bayes' rule, we can express the underlying semantic structure in the form of a posterior probability distribution. Since the posterior expression can not always be analytically computed, we estimate it using optimisation or direct sampling. Note that, in this project, we particularly focus on the topic inference using direct sampling. Relating to the attempts for applying the sampling-based inference to LDA-like models, the ground-work was established by Griffiths and Steyvers \cite{griffiths2004finding}. The authors display an application of the collapsed Gibbs sampling -- a Markov chain Monte Carlo (MCMC) algorithm. Since the method integrates the uncertainty out, we can sample the data entities without the explicit notion of the underlying parameters.

% Assumptions
% - Exchengeability (BoW)
% - Discrete data
\par Going back to the initial LDA model, the model's authors report on setting the following assumptions: exchangeability among the inner components of the lower and middle data hierarchy levels; and a discrete treatment of the lower-level data. By exchangeability, it is meant that the components follow the bag-of-words principle; that is, the order of the data has no correlation with the underlying semantic structure. Relating to the rationale behind the discrete data treatment, it is assumed that the lower-level components have no spatial connection. With respect to our project, both assumptions -- the exchangeability and the discreteness -- do not correspond to the characteristics of the metabolomics environment. Therefore, we establish a methodology to relaxing the latter assumptions.

@@

% Terminology intro:
% - Introduce the three-level hierarchical structure
\par In this subsection, we introduce our topic modelling terminology. In order to put the MSI data structure in the topic modelling context, the components of LDA's three-level hierarchical structure are defined as follows: \textit{a corpus} is the whole image of a sample; \textit{a document} is an MSI image's pixel; and \textit{a word} is an interval of mass-to-charge values corresponding to a particular ion. Note that from now on, we use the latter topic modelling concepts to introduce the LDA model's architecture and notation.

% LDA design
% - Graphical model
\par Since LDA-like models correspond to the branch of graphical models, the dependence of random variables can be illustrated using graphs. In order to familiarise with the architecture and the variables of LDA-like models, we provide Figure \ref{fig:arch_LDA-init} illustrating the initial LDA model in the plate notation.
\begin{figure}[h]
  \centering
  \includegraphics[width=0.20\textwidth]{LDA-initial.png}
  \caption{The initial LDA model's architecture.}
  \label{fig:arch_LDA-init}
\end{figure}
To start with, the circles indicate the model's variables: the coloured circle corresponds to the observable variable; and the uncoloured circles correspond to the hidden variables. Effectively, the hidden variables define the model's underlying structure. To introduce the plates, the plate denoted by $K$ corresponds to the number of topics; the plate denoted by $T$ corresponds to the number of documents, and the plate denoted by $N$ corresponds to the number of words. Effectively, the letters in the bottom right corners indicate the total number of variables. Therefore, a corpus has a $T$ number of documents, and each document has an $N$ number of words.

% Terminology
% - Parameter meaning
% - Listing
\par Before providing a listing with the definitions of the LDA variables, we introduce their purpose. The variables denoted by $\theta$ and $\phi$ correspond to the underlying probability distributions (e.g., in the case of the initial LDA model, we use Dirichlet distributions). It follows that the variables denoted by $\alpha$ and $\beta$ act as the parameters of the latter probability distributions; note that in the context of machine learning, such auxiliary parameters are called hyper-parameters. Effectively, \textit{a hyper-parameter} allows tuning a machine learning model for a particular dataset application. As a side reference, note that by \textit{a vocabulary} we refer to a collection of terms reflecting a fixed range of words. At this point, we provide the following list containing the definitions of the initial LDA model's variables: 
\begin{itemize}
	\item $K$ is the number of topics;
	\item $T$ is the number of documents;
	\item $N$ is the number of words per document;
	\item $V$ is the size of a vocabulary;
	\item $w$ is a word;
	\item $z$ is a word's topic assignment;
	\item $\theta_t$ is the topic distribution over the document $t$;
	\item $\phi_k$ is the vocabulary term distribution over the topic $k$;
	\item $\alpha$ is the hyper-parameter for the topic distributions;
	\item $\beta$ is the hyper-parameter for the vocabulary term distributions.
\end{itemize}

@@

% MSI dataset intro:
% - Introduce the raw data format
% - Describe the data pre-processing requirements
\par In this subsection, we introduce the qualities of the MSI data. Furthermore, we set the requirements for the pre-processing of raw MSI data; note that the pre-processing serves as an auxiliary method making raw MSI data compatible for a scalable topic modelling application. As a side note, the basis of our applied MSI data characteristics is established from the mzML data format. Conveniently, we parse mzML data using the \texttt{pymzml} Python library.

% Raw data format
% - Data meaning
% - Explain the raw data by looking into it
\par Recall that raw MSI data is mass spectra of a tissue's sample (i.e., the sample's every pixel is expressed in the form of a mass spectrum); also, every mass spectrum term conveys a particular intensity value. To help organise the data, we establish a continuous notion by ordering the mass-to-charge values. Further, since the sampling equipment can detect the mass-to-charge values in the millidalton (\textit{mDa}) precision, the MSI data is sparse (i.e., a large portion of mass-to-charge values are mapped to the zero intensity).

% Data pre-processing requirements
% - Vocabulary generation
% - Intensity normalisation
% - Dismissing inferior information
\par In the context of topic modelling, raw MSI data possess a large vocabulary (above $5000$ terms). Note that we consider every mass-to-charge value as a word; whereas every intensity value is perceived as a word's occurrence count. Further, since the intensity values could spike up above $1000$, the time complexity of the latent topic inference would require pro-longed runs. To overcome the introduced scalability issues, we carry data pre-processing (the applied techniques are discussed in the upcoming methodology section).

@@

% Contributions
% - Naive AR model application
% - Motivating for the state-of-the-art AR model application
\par The impact of proving the hypothesis would bring the following contributions:
\begin{itemize}
	\item Improve the detection of overlapping topics and vocabulary terms;
	\item Reduce the noisiness of the MSI data;
	\item Motivate a further research in the spatial smoothing application in MSI data.
\end{itemize} 
To expand on the overlapping topic detection, the issue arises when two underlying topics are made of similar vocabulary terms: instead of a separate representation, the topics are merged. We intend to identify the flow of distinct topics by applying spatial smoothing; as a consequence, the spatial smoothing would also impact the data noisiness reduction. Ultimately, if a naive spatial smoothing application displayed a performance improvement, the contribution would set a basis to utilise state-of-the-art auto-regression techniques in MSI data.

% Approach
% - Clearly define the experiment settings
% - Tune the models by studying synthetic datasets.
\par To my knowledge, the impact of applying spatial smoothing to MSI data has not yet been thoroughly studied. For the latter reason, this research project serves as an exploratory assessment -- we introduce the rationale behind the applied methodology; also, we clearly define the range of the experiment settings. To give a brief intuition about the methodology, the study assesses the domain-specific parameter tuning and its impact on a range of synthetic datasets.

@@

% Intro (separate into three)
% - MSI DATA:
% -- alonso2015analytical
% -- Review the metabolomics feature pre-processing
% -- Review the metabolomics alternative approaches
% - Review the implementation approaches
% -- blei2012probabilistic
% -- Gibbs sampling
% -- DTM
\par In this section, we review the ground-work carried on the following aspects: the pre-processing of MSI data; the rationale of the utilised topic models; and novel approaches exploiting the characteristics of MSI data. Note that the covered groundwork is selected to reflect the rationale of the utilised techniques. In order to establish the basis of the state-of-the-art computational metabolomics methodology, we consult the survey carried by Alonso et al. \cite{alonso2015analytical}; to introduce the key probabilistic topic modelling branches, we consult the survey carried by Blei \cite{blei2012probabilistic}.

@@

\par In order to establish a scalable topic inference, we review the following MSI data pre-processing techniques: data normalisation; feature binning; and noise reduction. To start with, data normalisation serves as a method establishing an adaptable data structure. Bolstad et al. \cite{bolstad2003comparison} have proposed the data normalisation method called linear baseline scaling. Note that the method is particularly targeted at sparse datasets. Effectively, the method is applicable to numerical features; note that the method's work-flow is carried as follows: we find the largest numerical feature of all data instances; then, we calculate the scaling factors by aligning the largest values to a pre-set upper threshold; finally, we align the remaining numerical features based on the established scaling factors. Relating linear baseline scaling to applications on MSI data, Kohl et al. \cite{kohl2012state} have shown that the method's application does not produce a significant loss of information to establish a well-performing MSI data inference.

% - Binning
% -- Evenly spaced 
% -- craig2006scaling
% -- Spectral peaks
% -- de2008nmr
\par To introduce feature binning, the technique is used to merge the instances of a feature space; as a result, feature binning reduces the data dimensionality. Note that by the MSI data feature space, we refer to distinct $m/z$ values. Since raw MSI data is sparse, a successful application of feature binning is based on identifying appropriate $m/z$ boundaries reflecting unique metabolite types. To introduce some examples, the research carried by Craig et al. \cite{craig2006scaling} have utilised an equally spaced feature binning. Even though the authors have succeeded to reduce the data dimensionality, they have encountered a loss of information by splitting the metabolite topics into arbitrary regions. As an alternative approach, De et al. \cite{de2008nmr} have performed a feature binning based on the identification of spectral peaks. Effectively, the technique induces a dynamic notion of bin boundaries which are based on the identified intensity peak regions. As a result, the merged $m/z$ values serve as a better representation of metabolite terms.

% - Denoising (removing small intensities)
% -- smith2014proteomics
\par The last reviewed MSI data pre-processing techniques addresses the MSI data noisiness. Even though the MSI data noise reduction is an open research problem utilising techniques beyond data pre-processing, Smith et al. \cite{smith2014proteomics} have shown that the application of a general data pre-processing routine displays performance improvements. For example, the authors have induced a lower intensity bound. Effectively, the intensity values below the intensity threshold would be treated as insignificant and/or as a product of data capturing device imperfections. By applying this procedure, the authors have successfully reduced the data dimensionality and, thus, increased the data scalability.

@@

% Intro
% - Inference
% - Derivatives
% - Gibbs sampling
% -- griffiths2004finding
\par In this subsection, we review the key characteristics of the prospective topic modelling inference techniques and model variations. To apply the inference techniques, we employ the rationale of a sampling-based LDA model. Griffiths and Steyvers \cite{griffiths2004finding} have displayed a successful application of the collapsed Gibbs sampler for the topic inference of both textual and visual data. To expand on the collapsed sampling, the technique allows skipping the estimation of the $\theta$ and $\phi$ values; instead, the inference is based on the notion of the assignment counts. In order to estimate a topic assignment's probability, the authors suggest using the following expression:
\begin{equation*}
P(z_i = k | z_{-i}, w) \propto \dfrac{n_{-i, k}^{(w_i)} + \beta}{n_{-i, k}^{(\cdot)} + V\beta}\cdot \dfrac{n_{-i, k}^{(t_i)} + \alpha}{n_{-i, \cdot}^{(t_i)} + K\alpha}.
\end{equation*}
Note that $i$ refers to the current iteration ($-i$ refers to the previous iteration); $k$ refers to a particular topic; and $n$ refers to the count of the instances indicated by the term's superscript. If required, $\theta$ and $\phi$ can be sampled from the following distributions:
\begin{align*}
\phi_{k,w} & \sim \mbox{Dirichlet}\bigg(\dfrac{n_{k}^{(w)} + \beta}{n_{k}^{(\cdot)} + W\beta}\bigg),\\
\theta_{t,k} & \sim \mbox{Dirichlet}\bigg(\dfrac{n_{k}^{(t)} + \alpha}{n^{(t)} + K\alpha}\bigg).
\end{align*}
Essentially, more accurate representations of $\theta$ and $\phi$ are obtained by running the Gibbs sampler until a sufficient convergence: we preserve the $\theta$ and $\phi$ values of each iteration; when the sampling is finished, the average of the preserved values is an accurate approximation of the underlying $\theta$ and $\phi$ values.

% Topic modelling:
% - DTM
% -- blei2006dynamic
% - SeqLDA
% -- du2012sequential
\par In order to relax the topic modelling assumptions of the initial LDA model, we review the model's derivatives. One of the assumptions -- word exchangeability -- is addressed by Blei and Lafferty \cite{blei2006dynamic}. 
The authors have proposed dynamic topic model (DTM) inducing the notion of change in the topic and vocabulary distributions; the model's architecture is illustrated in Figure \ref{fig:arch_dtm}.
\begin{figure}[h]
  \centering
  \includegraphics[width=0.4\textwidth]{dtm-architecture.png}
  \caption{The DTM architecture.}
  \label{fig:arch_dtm}
\end{figure}
Ultimately, the documents of a corpus are assigned to segments with unique $\theta$ and $\phi$ values. Note that the changes in the $\theta$ and $\phi$ values are impacted by their hyper-parameter updates. Since, in the DTM paper, the authors use variational inference, we provide a reference for an alternative dynamic topic modelling variation -- sequential LDA -- proposed by Du et al. \cite{du2012sequential}. Note that in the sequential LDA paper, the authors display a rigorous application of the collapsed Gibbs sampler.
